Paper 1: Instrument development

### 2024-09-02 Rmd/Paper1.Rmd
Added checks of linearity

library(readODS)                            # read .ods-files. Kilde: https://github.com/ropensci/readODS/
library(readxl)                             # read .xlsx-files. Kilde: https://readxl.tidyverse.org/
library(dplyr)                              # rename columns

@@

* data_TPACK_pilot, which uses TK, CK, etc for column names

@@

@@

Factor analysis is based on linear analysis (i.e., ordinary least squares-regression (OLS)). Assumptions underlying OLS-regression are according to @christophersen_introduksjon_2018 that the residuals have both an average value of 0, constant variance (i.e,. homoscedasticity) and independence (i.e., no autocorrelation). In addition, the dependent variables should depend linearly on the independent variables, which should have no multicollinearity (i.e., meaning that no residuals(?) of independent variable should be strongly correlated with any of the other independent variables). 
In addition, according to @flora_old_2012, it is important to decide whether to treat the data as ordinal or interval, as this has consequences for how the items are screened. Treating the data as interval, requires checking of missing data, linearity (checking for outliers and collinearity, i.e., very strong correlation between variables), and normally distributed residuals, checks which are not necessary to do if the data are treated as ordinal.

@@

@@

>>>>

library(readODS)                            # read .ods-files. Kilde: https://github.com/ropensci/readODS/
library(readxl)                             # read .xlsx-files. Kilde: https://readxl.tidyverse.org/

library(pander)                             # print pretty tables
library(dplyr)                              # rename columns
library(tidyr)                              # pivot_longer

library(ggplot2)

# multicollinearity checks
library("corrplot")                         # multicollinearity 
library(car)                                # vif                      
library("olsrr")                            # vif

@@

* data_TPACK_pilot, which uses TK, CK, etc for column names
* data_TRELIS_Q

data_TPACK_pilot will be used for the EFA, while data_TRELIS_Q will be used for the CFA.

@@

In the following we will only consider the pre-datasett, so we remove the post-data, which is to dissimilar from the pilot data.

```{r select_data_for_screening}
# emne2_anon (uten post)
data_TPACK_pilot_minuspost <- data_TPACK_pilot[
  data_TPACK_pilot$datasett != "post" & 
  data_TPACK_pilot$datasett != "pilot_trelis_q",]
```

@@

Factor analysis is based on linear analysis (i.e., ordinary least squares-regression (OLS)). Assumptions underlying OLS-regression are according to @christophersen_introduksjon_2018 that the residuals have both an average value of 0, constant variance (i.e,. homoscedasticity) and independence (i.e., no autocorrelation). In addition, the dependent variables should depend linearly on the independent variables, which should have no multicollinearity (i.e., meaning that no residuals(?) of independent variable should be strongly correlated with any of the other independent variables). 
In addition, according to @flora_old_2012, it is important to decide whether to treat the data as ordinal or interval, as this has consequences for how the items are screened. Treating the data as interval, requires checking of missing data, linearity (checking for outliers and collinearity, i.e., very strong correlation between variables), and normally distributed residuals, checks which are not necessary to do if the data are treated as ordinal. In addition, checks suggested by @izquierdo_exploratory_2014 were done, such as check of eigenvalues, item-test correlation and reliabilities (i.e., Cronbach's alpha).

@@

A list of items is compiled for easier retrieval.

```{r}
TK_pos <- c("TK1","TK2","TK3","TK4","TK5","TK6","TK7","TK8")
PK_pos <- c("PK1","PK2","PK3","PK4","PK5","PK6")
CK_pos <- c("CK1","CK2","CK3","CK4","CK5","CK123","CK6")
PCK_pos <- c("PCK1","PCK2","PCK3","PCK4","PCK5")
TCK_pos <- c("TCK1","TCK2","TCK3","TCK4")
TPK_pos <- c("TPK1","TPK2","TPK3","TPK4","TPK5")
TPACK_pos <- c("TPACK1","TPACK2","TPACK3","TPACK4","TPACK5")

#TPACK_complete <- c(TK_pos, PCK_full_pos, TPACK_full_pos) # minus XK

TPACK_list <- list(TK=TK_pos,
                   CK=CK_pos,
                   PK=PK_pos,
                   PCK=PCK_pos,
                   TCK=TCK_pos,
                   TPK=TPK_pos,
                   TPACK=TPACK_pos)

# CK6 and CK123 is only used for the TRELIS-Q dataset, not the pilot dataset
TPACK_list$CK <- TPACK_list$CK[!TPACK_list$CK %in% c("CK6","CK123")]
```

@@

### Linearity?

Here I do checks for linearity, collinearity and multicollinearity.

Through an iterative process, the following items were identified as contributing to high collinearity (r>0.9). They were therefore excluded from the dataset.

```{r remove_collinear_items}
data_TPACK_pilot_minuspost <- data_TPACK_pilot_minuspost %>%
    # 25.5.2023: velg bort items som ikke er i TRELIS-Q
    select(-c("TK8","PK2","PK4","PK5","PK6","PCK1","PCK3","PCK4","TCK2","TCK3",
              "TPK2","TPK3","TPK5","TPACK1","TPACK5",
              "TPACK3", # colinear in trelis-q; "TK1","TPK1",
              "TK6","TK4", "PK1", "PK3"))
```

```{r linear_scatterplots}
# scatterplot som korrigerer for overlap mellom datapunkter
teller <- 0                          # hvor mange plot lager jeg?
datasett_index <- 1
start <- 1                          # siste item = datasett

#for (dataset in names(data_list)) {
  #datasett <- data_TPACK_pilot_minuspost
  end <- ncol(data_TPACK_pilot_minuspost)
  i <- 1                               # start posisjon
  for (var1 in names(TPACK_list)) {           # TPACK-element
    items <- colnames(data_TPACK_pilot_minuspost)[colnames(data_TPACK_pilot_minuspost) %in% TPACK_list[[var1]]]
    ifelse (length(items)>1, 
      y <- rowMeans(data_TPACK_pilot_minuspost[, items]), # y-variabel
      y <- data_TPACK_pilot_minuspost[, items])           # y består av bare ett item
    i <- i + 1                         # for å unngå duplikater
    for (var2 in items) {               # x-variabel
      # scatterplot
      # kilde: https://ggplot2-book.org/statistical-summaries.html
      # kilde: https://stackoverflow.com/questions/32004292/r-ggplot-for-loop-aes-variable
      x <- data_TPACK_pilot_minuspost[,var2]
      xy <- data.frame(t(rbind(y,x)))
      colnames(xy) <- c(var1,var2)

      # check for collinearity, r>0.85 / 0.90
      #print("collinearity: r > 0.85?")
      r <- cor.test(x,y)
      #print(r)   # marker verdier større enn 0.85

      # 0.4 er moderat korrelasjon ift Christophersen (2018) kap. 3
      if (abs(r$estimate)>0.4) { # only plot scatterplots which may be linear?
        xy_plot <- ggplot(xy, aes_string(var2,var1)) +
          geom_point() +                            # shape=1,alpha=1/5
          geom_jitter() +
          ggtitle(paste("scatterplot", data_TPACK_pilot_minuspost$datasett))

        print(xy_plot)
        teller <- teller + 1   # count number of scatterplots
      }
    }
  }
#  datasett_index <- datasett_index + 1
#}
print(teller) # antall plot
```

```{r linear_scatterplots_aggregated}
teller <- 0                          # hvor mange plot lager jeg?
datasett_index <- 1
start <- 1                          # siste item = datasett

#for (dataset in names(data_list)) {
  #datasett <- data_TPACK_pilot_minuspost
  end <- ncol(data_TPACK_pilot_minuspost)
  i <- 1                               # start posisjon
  for (var1 in names(TPACK_list)) {           # TPACK-element
    items <- colnames(data_TPACK_pilot_minuspost)[colnames(data_TPACK_pilot_minuspost) %in% TPACK_list[[var1]]]

    if (length(items) == 0) {next} # skip empty items

    ifelse (length(items)>1,  # calculate average value
      y <- rowMeans(data_TPACK_pilot_minuspost[, items]), # y-variabel
      y <- data_TPACK_pilot_minuspost[, items])

    x <- data_TPACK_pilot_minuspost[, items]

    xy <- data.frame(y,x)
    colnames(xy) <- c(var1,items)

    # pivot_longer
    # Warning: Using an external vector in selections was deprecated in tidyselect 1.1.0.
    # ℹ Please use `all_of()` or `any_of()` instead.
    xy_long <- pivot_longer(xy,!c(all_of(var1)), names_to="items",values_to = "likert")

    #Warning: `aes_string()` was deprecated in ggplot2 3.0.0.
    #ℹ Please use tidy evaluation idioms with `aes()`.
    #ℹ See also `vignette("ggplot2-in-packages")` for more information.

       xy_plot <- ggplot(xy_long, aes_string("likert",var1)) +
          geom_point() +                            # shape=1,alpha=1/5
          geom_jitter() +
          ggtitle(paste("scatterplot", data_TPACK_pilot_minuspost$datasett))

        print(xy_plot)
        teller <- teller + 1   # count number of scatterplots
    }
  #datasett_index <- datasett_index + 1
  #}
print(teller) # antall plot
```

```{r check_collinearity}

### HELP FUNCTIONS

lag_eigenvalue_tabell <- function(ev_list) {
  no_ev <- length(ev_list) # antall eigenvalues
  kum_var <- 0

  # gjør om til tabell
  ev_tabell <- data.frame(matrix(ncol=4,nrow=0))
  for (i in 1:no_ev) {
    ev <- ev_list[i]
    kum_var <- kum_var + ev/no_ev
    faktor <- cbind(as.integer(i), ev, ev/no_ev, kum_var)
    #print(faktor)
    ev_tabell <- rbind(ev_tabell, faktor)
  }

  colnames(ev_tabell) <- c("Factor","Eigenvalue","Variance","Cummulativ variance")

  pander(format(ev_tabell,digits=2))
}

collinear <- function (dataset) {
  ###
  # Christophersen (2018): r > 0.9?
  ###

  teller <- 0                          # hvor mange sjekker gjør jeg?
  start <- 1                           
  end <- ncol(dataset)
  i <- 1                               # start posisjon
  for (var1 in colnames(dataset[start:(end-2)])) {
    i <- i + 1                         # for å unngå duplikater
    for (var2 in colnames(dataset[i:(end-1)])) {
      # check for collinearity, r>0.90
      r <- cor.test(dataset[,var1], dataset[,var2])

      if (abs(r$estimate)>0.9) { 
        print("collinearity: r > 0.90")
        print(c(dataset, var1 ,var2, r$estimate))   # marker verdier større enn 0.90

        teller <- teller + 1   # count number of scatterplots
      }
    }
  }

  if (teller == 0) {
    print("Ingen kolinearitet oppdaget") # antall plot
  }

  ###
  # Flora et al. (2012): eigenvalues <= 0? condition index < 30?
  ###
  ev <- eigen(cor(dataset,use="pairwise.complete.obs"), only.values=TRUE)
  ev_table <- lag_eigenvalue_tabell (ev$values)
  #ev
  # beregn condition index: ci < 30
  ci <- ev$values[1]**2 - ev$values[length(ev$values)]**2 
  print(c("Condition index(<30?):", ci))
  return(ev_table)
}

#datasett <- data_list[[1]]
# eigenvalue_table
ev_table <- collinear(data_TPACK_pilot_minuspost[,colnames(data_TPACK_pilot_minuspost) %in% unlist(TPACK_list)])
ev_table
```

```{r check_multicollinearity}
# multiple regression model

# I browse through the items, and remove the current item by subtracting it at the end of the formula
#for (dataset in names(data_list)) {
  mc <- NULL      # list of items with multicollinearity

      # kilde: https://www.codingprof.com/3-ways-to-test-for-multicollinearity-in-r-examples/
      corrplot(cor(data_TPACK_pilot_minuspost[,colnames(data_TPACK_pilot_minuspost) %in% unlist(TPACK_list)]), method = "number", number.cex=0.3) # r bør være under 0.85

      # forenkle navn på dataset for prettyprint
      df <- data_TPACK_pilot_minuspost

      #items <- unlist(TPACK_list)
      items <- colnames(df[,colnames(df) %in% unlist(TPACK_list)]) # ikke ha med subset-kolonnen

      # lag tekststreng for uavhengige variable
      #antall_tpack <- length(grep("TPACK",items))
      independent_var <- paste("'] ~ df [,'", items[1], sep="") #TK1"
      for (item in items[2:(length(items))]) { #-antall_tpack)]) { # omit TPACK
        independent_var <- paste(independent_var, "'] + df [,'", item, sep="")
      }
      independent_var <- paste(independent_var, "'] - df  [,'", sep="")

      for (item in items) {
        model <- lm( as.formula(
        paste( "df [,'", item, independent_var, item, "']", sep="")))

        print("############################")
        print("######## NEW ITEM ##########")
        print("############################")

        print(item)

        # check for multicollinearity: R > 0.9
        print("Any R>0.9?")
        mc_items <- model$coefficients[abs(model$coefficients)>0.9]
        if (length(mc_items) > 0) {
          print(c("multicollinearitet R > 0.9?", mc_items))
          print(summary(model))
          mc <- c(mc, item)
        }

        #if (vif == 1) {
          # kilde: https://www.statology.org/multicollinearity-in-r/
          print("car::vif")
          print(vif(model))

          print("olsrr::vif")
          # kilde: https://www.codingprof.com/3-ways-to-test-for-multicollinearity-in-r-examples/
          print("tolerance bør være over 0.1 og VIF under 10.")
          print(ols_vif_tol(model)) # tolerance bør være over 0.1 og VIF under 10.
          #print("CI bør både være under 10 og varians lav")
          #print(ols_eigen_cindex(model)) # CI bør både være under 10 og varians lav
        #}
      }

# print items with multicollinearity
print(mc)
#}
```

No multicollinearity. The items added to mc are due to the intercepts, not the slopes.
